---
title: "Analyzing FRET binding data"
author: "Guillaume Gaullier"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        toc: yes
        fig_width: 5
        fig_height: 3.76
vignette: >
  %\VignetteIndexEntry{Using rfret}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This documentation explains how to use the functions provided by the `rfret`
package to analyze FRET binding data from titration experiments, and determine
the binding constant (Kd) of a macromolecular interaction. An example using a
real-life dataset is presented and discussed.

# Overview
## FRET data analysis workflow

Analysis of FRET data can be broken down into the following steps, each of
which `rfret` provides functions for:

1. inspect raw data to detect possible outliers (function
   `fret_inspect_raw_data`);
2. average fluorescence intensities across technical replicates, if such
   replicates were measured (function `fret_average_replicates`);
3. correct FRET signal of the titration experiment, using signal from blank
   experiments which contain no donor or no acceptor fluorophores (function
   `fret_correct_signal`);
5. fit a binding model equation to the experimental data, to determine the
   binding constant Kd (function `fit_binding_model`).

The `fit_binding_model` function is independent of the others. It can accept any
data from other types of experiments, as long as this data comes as two columns
called `concentration` and `signal`.

On the other hand, all the FRET data processing functions (named with a common
`fret_` prefix) are designed to work together: each function produces an output
that can be directly used as input for the next one, making the whole analysis
free of any need for more data reformatting as long as input files are properly
formatted in the first place. These functions work ideally with the pipe
operator (`%>%`) from [the `magrittr` package][magrittr].

## Raw data format

`rfret` is designed such that the user doesn't have to manually do any
calculation or any edit to the numbers obtained from a plate reader instrument:
such manipulations in spreadsheet programs are time-consuming and error-prone,
and doing calculations in a spreadsheet can accidentally alter the raw data.

The raw data must be formatted like the following example (first and last 6 rows
of a real-life dataset):
```{r Raw data format (head), echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(head(rfret::fret_binding_data))
```

```{r Raw data format (tail), echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(tail(rfret::fret_binding_data))
```

Most (if not all) plate reader instruments have an "export to CSV" feature,
which is best suited to save raw data in a format usable by `rfret`. If offered
a choice between "plate layout" and "table view", choose "table view": this will
export the data in a format similar to the previous table.

The `Well Row` and `Well Col` columns present in this example dataset were
created by the plate reader software used to produce this dataset, but `rfret`
will simply ignore them and any other column except the following ones (these
can come in any order, but must have the exact following case sensitive names):

- `Content`
- `fret_channel`
- `acceptor_channel`
- `donor_channel`
- `concentration`

Therefore, you will need to either edit these column names manually when you
save your data, or set up your plate reader software to write these column
names automatically for you. You might need to add the `concentration` column
manually, if your plate reader software is not aware of your titration series.

The `Content` column should contain, for each row, a name indicating what part
of the experiment and what replicate this row belongs to, with the format
`type.replicate` where `type` can be any word, and `replicate` is a number. They
must be separated by a dot, and there can be only one dot in the entire resulting
character string. The example dataset contains the following contents:
```{r Content column levels, echo=FALSE, message=FALSE, warning=FALSE}
levels(as.factor(rfret::fret_binding_data$Content))
```

This data preparation step is the only one for which you will need to use a
spreadsheet program (you can also edit the CSV file in a simple text editor,
but a spreadsheet displays this sort of data in a way that is much easier to
navigate). This step can be automated in the plate reader software as well.

# Analyzing a dataset

We first need to load the `rfret` and `magrittr` packages:
```{r Load packages}
library(magrittr)
library(rfret)
```

## Data preparation for processing

All `fret_` functions in this package work by first agregating together any
number of datasets, to perform all subsequent data processing and analysis steps
on a single, large internal data table. Any analysis must start by using the
`fret_format_data` function to prepare the raw data.

The function `fret_format_data` takes either:

- the name of a single directory from which it will read all CSV files,
- a single file name or a set of file names,
- a single dataset already loaded in memory as a dataframe,
- or a list of datasets already loaded in memory as dataframes.

It will format the data properly for subsequent processing by the other
functions.

## Raw data inspection

A FRET experiment provides built-in controls for the correct preparation of the
titration samples. In a typical experiment, the donor-labeled molecule is kept
constant at a low concentration, and the acceptor-labeled molecule is titrated
across a wide concentration series prepared by serial dilutions. Fluorescence
detection is sensitive enough to detect very small pipetting errors, therefore,
inspecting the fluorescence intensities in the donor and acceptor channels 
provides an easy way to check for outliers in the series.

The function `fret_inspect_raw_data` takes the output of `fret_format_data` and
outputs, for each experiment, plots of the fluorescence intensity across the
titration series for the donor, acceptor and FRET channels.

Let's look at an example:
```{r Inspect raw data, message=FALSE, warning=FALSE}
fret_binding_data %>%
    fret_format_data() %>%
    fret_inspect_raw_data()
```

In the donor channel plot, all fluorescence intensities should be equal, since
the donor-labeled molecule is present at the same concentration in all samples.
In the titration experiment, it is expected that the fluorescence intensity of
the donor channel will decrease at high concentration of acceptor if FRET happens
(because part of the donor fluorescence is transferred to the acceptor instead of
being emitted).

In the acceptor channel plot, the fluorescence intensity should increase linearly
with the acceptor-labeled molecule concentration across the titrations series.
If the serial dilution was performed with good precision, all data points should
cluster around a straight line, as in this example.

The FRET channel plot quickly tells if the data contains useful signal: the
titration and acceptor_only experiments should have visible differences in
fluorescence intensity at some point in the titration series. If their data
points perfectly overlap, it means the data doesn't contain any useful signal.

## Data processing

Automatic processing of raw data can be achieved using the pipe operator `%>%`
from [the `magrittr` package][magrittr], and the three functions
`fret_format_data`, `fret_average_replicates` and `fret_correct_signal`:
```
library(magrittr)
fret_binding_data %>%
    fret_format_data() %>%
    fret_average_replicates() %>%
    fret_correct_signal(output_directory = "./my_results")
```

The pipe operator allows to write this code in a natural way, one step after the
other in the order of execution.

### Average technical replicates

The function `fret_average_replicates` takes a raw dataset from
`fret_format_data` and returns a dataset where the fluorescence intensity values
have been averaged over the technical replicates.

### Correct FRET signal

The function `fret_correct_signal` applies the correction described in the
following references:

- Hieb AR *et al* (2012) Fluorescence Strategies for High-Throughput
  Quantification of Protein Interactions. *Nucleic Acids Research* 40 (5): e33
  <https://doi.org/10.1093/nar/gkr1045>
- Winkler DD *et al* (2012) Quantifying Chromatin-Associated Interactions: The
  HI-FI System. In *Methods in Enzymology* pp 243–274. Elsevier.
  <https://doi.org/10.1016/B978-0-12-391940-3.00011-1>

It takes a reduced dataset from `fret_average_replicates` and outputs the
corrected FRET signal. Given a directory name, it will write out the corrected
datasets to CSV files. These output files will have the same names as the input
files, appended with `_corrected`. These files can be used to plot the
corrected data using a different software, for example.

## Curve fitting

The last step involves the function `fit_binding_model`. It takes a corrected
dataset (output of `fret_correct_signal`) and a binding model. Currently available
binding models are: `hyperbolic`, `hill` and `quadratic`. To use the quadratic,
model the function also need the fixed concentration of probe molecule. It
returns an `nls` object that can be further examined with standard functions
(like `summary`) or used as input for `make_figure`.

Fitting the example dataset with the quadratic model can be achieved as follows:
```{r Curve fitting, message=FALSE, warning=FALSE}
fret_binding_data %>%
    fret_format_data() %>%
    fret_average_replicates() %>%
    fret_correct_signal() %>%
    fit_binding_model(binding_model = "quadratic",
                      probe_concentration = 5)
```

Given a directory name, `fit_binding_model` will write its results to CSV files.

## Final figure

The function `make_figure` takes the output of `fit_binding_model` and returns
a figure with a plot of the binding curve and a residual plot. Given a directory
name, it will write the figure in a file named like the original input file.
The figure is written as a PNG file by default, but the function can also write
PDF and SVG files.
```{r Figure making, message=FALSE, warning=FALSE}
fret_binding_data %>%
    fret_format_data() %>%
    fret_average_replicates() %>%
    fret_correct_signal() %>%
    fit_binding_model(binding_model = "quadratic",
                      probe_concentration = 5) %>%
    make_figure(probe_concentration = 5)
```

# Automated processing and analysis

The entire data inspection, processing and analysis can be performed by a single
script run from the directory containing raw data files in CSV format. Here is
an example script:
```
# Load required packages
library(magrittr)
library(rfret)

# List data files
my_files <- list.files(path = ".", pattern = ".csv")

# Generate raw data plots and save them in a directory called 'data_inspection'
my_files %>%
    fret_format_data(skip_lines = 4) %>%
    fret_inspect_raw_data(output_directory = "./data_inspection",
                          plot_format = "png")

# The following commands will:
# 1. process raw data and save corrected data in CSV files in a
# directory 'corrected_data'
# 2. fit each dataset with a quadratic model equation, and save fit results
# in CSV files in a directory 'fit_results'
# 3. generate binding curve figures and save them as PNG files in a
# directory 'final_figures'
my_files %>%
    fret_format_data(skip_lines = 4) %>%
    fret_average_replicates() %>%
    fret_correct_signal(output_directory = "./corrected_data") %>%
    fit_binding_model(binding_model = "quadratic",
                      probe_concentration = 5,
                      output_directory = "./fit_results") %>%
    make_figure(probe_concentration = 5,
                output_directory = "./final_figures",
                plot_format = "png")
```

# Automated report generation

Used in combination with [R Markdown][rmarkdown], `rfret` can make the process of
data analysis fully integrated with lab notebook record keeping.

[magrittr]: http://magrittr.tidyverse.org
[rmarkdown]: http://rmarkdown.rstudio.com
