---
title: "Using rfret"
author: "Guillaume Gaullier"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        fig_width: 5
        fig_height: 3.76
vignette: >
  %\VignetteIndexEntry{Using rfret}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This documentation explains how to use the functions provided by the `rfret`
package to analyze FRET titration experiments and determine a macromolecular
interaction's binding constant (Kd). Examples using real-life datasets are
presented and discussed.

# Overview
## FRET data analysis workflow

Analysis of FRET data can be broken down into the following steps, each of
which `rfret` provides functions for:

1. inspect raw data to detect possible outliers (function `inspect_raw_data`);
2. average fluorescence intensities across technical replicates, if such
   replicates were measured (function `average_technical_replicates`);
3. correct FRET signal of the titration experiment, using signal from blank
   experiments which contain no donor and no acceptor (function
   `correct_fret_signal`);
4. guess initial values for the parameters of the quadratic binding equation
   (`kd`, `fret_min` and `fret_max`; function `guess_quadratic_parameters`);
5. fit the quadratic binding equation to the experimental data, to determine the
   binding constant Kd (function `fit_binding_model`).

All these functions are designed to work independently, therefore you can skip
some of these steps (for example, you might not have technical replicates to
average, or you might perform your own signal correction using a different
method and software). However, they are also designed to produce outputs that
can be directly used as input for the next function, making the whole analysis
free of any need for more data reformatting.

## Raw data format

`rfret` is designed such that the user doesn't have to manually do any
calculation or any edit to the numbers: such manipulations in spreadsheet
programs are not only time-consuming, they are also error-prone and encourage
the bad habit of doing calculations in the same file that contains the raw data
(which should remain unaltered, to ensure reproducibility of the analysis).

The raw data must be formatted like the following example (first 6 rows of a
real-life dataset):
```{r Raw data format, echo=FALSE}
knitr::kable(head(rfret::fret_good))
```

Most (if not all) plate readers have an "export to CSV" feature, which is best
suited to save raw data in a format usable by `rfret`. If offered a choice
between "plate layout" and "table view", choose "table view": this will export
the data in a format similar to the previous table.

The `Well.Row` and `Well.Col` columns present in this example dataset were
created by the plate reader software used to produce this dataset, but `rfret`
will simply ignore them and any other columns except the following ones (these
can come in any order, but must have the exact following case sensitive names):

- `Content`
- `fret_channel`
- `acceptor_channel`
- `donor_channel`
- `concentration`

Therefore, you will need to either edit these column names manually when you
save your data, or set up your plate reader software to write these column
names automatically for you. You might need to add the `concentration` column
manually, if your plate reader software is not aware of your titration series.

This data preparation step is the only one for which you will need to use a
spreadsheet program (you can also edit the CSV file in a simple text editor,
but a spreadsheet displays this sort of data in a way that is much easier to
navigate).

# Analyzing a dataset
## Inspect raw data

By its very nature, a FRET experiment provides built-in controls for the correct
preparation of the titration samples. In a typical experiment, the donor-labeled
molecule is kept constant at a low concentration, and the acceptor-labeled
molecule is titrated across a wide concentration series prepared by serial
dilutions. In these conditions, inspecting the fluorescence intensities in the
donor and acceptor channels provides an easy way to check that these
concentrations meet the expected ones.

To use `rfret`, we first need to load the package:
```{r Load rfret}
library(rfret)
```

The function `inspect_raw_data` takes a raw dataset and outputs, for each one of
the donor, acceptor and FRET channels, a plot of the fluorescence intensity
across the titration series. Warnings will appear, because two of these plots
have a logarithmic scale on the `x` axis, and some datapoints have zero as `x`
coordinate (samples with no acceptor-labeled molecule).

Let's look at an example:
```{r Inspect raw data}
ttn <- c("titration_1", "titration_2")
plots_1 <- inspect_raw_data(raw_data = fret_good, titrations = ttn)
plots_1$donor
```

In the donor channel plot, all fluorescence intensities should be equal, since
the donor-labeled molecule is present at the same concentration in all samples.
Fluorescence detection is sensitive enough to detect very small pipetting errors,
therefore this plot is a good way to detect such errors. The line represents the
mean of all fluorescence intensities in the donor channel, and the shaded area
represents +/- 10 % of the mean. If all data points fall into this shaded area,
as in this example (except one slightly outside), then the pipetting of the
donor-labeled molecule achieved a good precision.

The blank experiment doesn't contain the donor-labeled molecule, and has
near-zero fluorescence intensities across the entire titration series, as
expected.

Let's look at the acceptor channel plot:
```{r Acceptor plot}
plots_1$acceptor
```

If the serial dilution was performed with good precision, all data points should
cluster around a straight line, as in this example.

The FRET channel plot quickly tells if the data contains useful signal, and if
it is worth completing the analysis:
```{r FRET plot}
plots_1$fret
```

The titration and blank experiments should have visible differences in
fluorescence intensity at some point in the titration series. If their data
points perfectly overlap, it means the data doesn't contain any useful signal
(as we will see in a later example).

## Average technical replicates

The function `average_technical_replicates` takes a raw dataset, and the names
of the replicates associated to the titration and blank experiments (column
`Content` in the raw dataset). These names are chosen by the user at the data
saving stage right after the experiment. In our example dataset, these names are
`blank_1`, `blank_2`, `titration_1` and `titration_2`, and each one has 24 data
points:
```{r Technical replicates}
summary(fret_good$Content)
```

We already set the names of the titration experiments above, so we can reuse
them and only need to define the blank names. We will store the resulting
reduced dataset in a variable to use it in subsequent steps of the analysis.
```{r Average replicates}
blk <- c("blank_1", "blank_2")
reduced_dataset <- average_technical_replicates(raw_data = fret_good,
                                                titrations = ttn,
                                                blanks = blk)
```

The reduced datasets now have only two names, with 24 data points each:
```{r Merged replicates}
summary(reduced_dataset$Content)
```

## Correct FRET signal

The function `correct_fret_signal` applies the correction described in the
following references:

- Hieb AR *et al* (2012) Fluorescence Strategies for High-Throughput
  Quantification of Protein Interactions. *Nucleic Acids Research* 40 (5): e33
  <https://doi.org/10.1093/nar/gkr1045>
- Winkler DD *et al* (2012) Quantifying Chromatin-Associated Interactions: The
  HI-FI System. In *Methods in Enzymology* pp 243–274. Elsevier.
  <https://doi.org/10.1016/B978-0-12-391940-3.00011-1>

It takes a reduced dataset from `average_technical_replicates` (or any similarly
formatted dataset) and outputs the corrected FRET signal.
```{r Correct FRET signal}
final_dataset <- correct_fret_signal(reduced_dataset)
final_dataset
```

## Guess initial values of the binding model equation parameters

The function `guess_quadratic_parameters` takes a corrected dataset from
`correct_fret_signal` (or any similarly formatted dataset) and outputs initial
values for `kd`, `fret_min` and `fret_max`.

These values are based on very naive guesses: `fret_min` and `fret_max` are
simply the minimum and maximum fluorescence intensities present in the final
corrected dataset, and `kd` is the value of `concentration` of the datapoint
that has the closest FRET value to `(fret_max - fret_min) / 2` (fluorescence
intensity at half-maximum).

These naive guesses are most of the time close enough to the actual value to
make the fit converge properly.

Let's try that on our example dataset:
```{r Guess parameters}
guessed_params <- guess_quadratic_parameters(final_dataset)
guessed_params
```

## Curve fitting

The last step involves the function `fit_binding_model`, which is simply a
wrapper for R's `stats::nls` function. It has the quadratic binding equation
built-in. It takes a corrected dataset (output of `correct_fret_signal`), the
fixed concentration of donor-labeled molecule, and a list of initial values for
`kd`, `fret_min` and `fret_max` (either provided by the user, or directly the
output of `guess_quadratic_parameters`), and returns a result table, a plot of
the binding curve, a residual plot, and a figure with those two plots combined.

Let's finish analyzing our example dataset, display the result table and the
final figure:
```{r Fit binding model}
my_fit <- fit_binding_model(fret_corrected = final_dataset,
                            donor_concentration = 10,
                            parameters = guessed_params)
my_fit$results
my_fit$final_figure
```

# Detecting outliers

The previous example featured a good dataset, making the analysis
straightforward. The following two examples show how the function
`inspect_raw_data` can be useful to detect outliers.

## Donor outliers

Let's inspect a different dataset:
```{r Donor outliers}
plots_2 <- inspect_raw_data(fret_donor_outliers, ttn)
plots_2$donor
```

The donor channel plot clearly shows four obvious outliers in the `titration_1`
series. They are at the 5^th^, 13^th^, 21^st^ and 24^th^ positions in the
titration series. We can easily find them in the raw data:
```{r}
fret_donor_outliers$donor_channel[fret_donor_outliers$Content == "titration_1"][c(5, 13, 21, 24)]
```

These outliers come from pipetting errors: the three wells that have a higher
fluorescence intensity received too much donor-labeled molecule, whereas the
one that has a lower fluorescence intensity did not receive enough donor-labeled
molecule. Knowing that, one can decide on a rationale basis to exclude these
data points from the subsequent analysis.

## Acceptor outliers

Let's inspect another different dataset:
```{r Acceptor outliers}
plots_3 <- inspect_raw_data(fret_acceptor_outliers)
plots_3$acceptor
```

This acceptor channel plot shows that the highest acceptor-labeled molecule
concentration is very different between the titration and blank experiments, and
between technical replicates. Again, one can decide to exclude data points from
the analysis based on this observation.

# Detecting usable FRET signal

The previous two datasets are not ideal, because they contain outliers in either
the donor channel or the acceptor channel. Yet, they contain useful FRET signal
and can yield binding curves and binding constants. You now know how to analyze
them to find out what these binding curves look like, and what the associated
binding constants are.

In this last example, we will see how the function `inspect_raw_data` can help
quickly detecting usable signal in a FRET dataset. Let's inspect a last dataset:
```{r No FRET signal}
plots_4 <- inspect_raw_data(fret_bad)
plots_4$fret
```

Something looks wrong here: there is very little visible difference in
fluorescence intensity in the FRET channel between the titration and blank
experiments.

For comparison, this is how this same plot looked for the first dataset we
inspected:
```{r Detectable FRET signal}
plots_1$fret
```

The last example dataset doesn't look like it contains any signal. We can verify
that by plotting the corrected data and see if we get anything that looks like a
binding curve:
```{r Flat binding curve}
bad_data <- average_technical_replicates(fret_bad, ttn, blk)
bad_data_corr <- correct_fret_signal(bad_data)
library(ggplot2)
qplot(concentration, fret_corrected, data = bad_data_corr, log = "x") + theme_bw()
```

This dataset doesn't yield any binding curve, and trying to fit this data would
either fail or give a completely meaningless value for Kd.

# Automatic report generation

Used in combination with [R Markdown][rmarkdown], `rfret` makes the process of
data analysis fully integrated with the lab notebook record keeping process.

[rmarkdown]: http://rmarkdown.rstudio.com
